---
title: "Prostate Cancer Clustering"
subtitle: "Hierarchical, partitioning, and model-based methods"
author: "George Oikonomidis"
editor: visual
format:
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
---

## Libraries

```{r}
set.seed(125)

library(dplyr)
library(ggplot2)
library(patchwork)
library(cluster)
library(factoextra)
library(mclust)
library(reshape2)
```

## Data preparation

The diagnosis label is withheld from clustering and used only for external evaluation. All eight continuous measurements remain continuous.

```{r}
prostate <- read.csv("data/prostate_cancer.csv", header = TRUE)
prostate$diagnosis_result <- factor(
  prostate$diagnosis_result,
  levels = c("B", "M"),
  labels = c("Benign", "Malignant")
)

clustering_variables <- c(
  "radius", "texture", "perimeter", "area",
  "smoothness", "compactness", "symmetry",
  "fractal_dimension"
)

clustering_data <- prostate[clustering_variables]

for (variable in clustering_variables) {
  if (anyNA(clustering_data[[variable]])) {
    clustering_data[[variable]][is.na(clustering_data[[variable]])] <-
      median(clustering_data[[variable]], na.rm = TRUE)
  }
}

data.frame(
  observations = nrow(clustering_data),
  benign = sum(prostate$diagnosis_result == "Benign"),
  malignant = sum(prostate$diagnosis_result == "Malignant")
)

str(clustering_data)
```

## Exploratory analysis

```{r}
histograms <- lapply(clustering_variables, function(variable) {
  ggplot(clustering_data, aes(x = .data[[variable]])) +
    geom_histogram(color = "black", fill = "#8C87B4", bins = 30) +
    theme_minimal() +
    labs(title = variable, x = NULL, y = "Count")
})

wrap_plots(histograms, ncol = 2)
```

```{r}
boxplots <- lapply(clustering_variables, function(variable) {
  ggplot(clustering_data, aes(y = .data[[variable]])) +
    geom_boxplot(fill = "#8C87B4", outlier.colour = "red") +
    theme_minimal() +
    labs(title = variable, x = NULL, y = NULL)
})

wrap_plots(boxplots, ncol = 4)
```

The IQR rule is used to identify potential outliers, not to replace them with column means.

```{r}
outlier_summary <- lapply(clustering_variables, function(variable) {
  values <- clustering_data[[variable]]
  first_quartile <- quantile(values, 0.25)
  third_quartile <- quantile(values, 0.75)
  interquartile_range <- third_quartile - first_quartile
  lower_limit <- first_quartile - 1.5 * interquartile_range
  upper_limit <- third_quartile + 1.5 * interquartile_range
  outliers <- values[values < lower_limit | values > upper_limit]

  data.frame(
    Variable = variable,
    Number_of_outliers = length(outliers),
    Outlier_values = paste(outliers, collapse = ", ")
  )
})

bind_rows(outlier_summary)
```

```{r}
original_scale_data <- clustering_data
standardised_data <- scale(clustering_data)
```

## Hierarchical clustering with Manhattan distance

Single, complete, and average linkage are retained for Manhattan distance. Ward clustering is deliberately excluded because its minimum-variance interpretation requires squared Euclidean geometry.

```{r}
manhattan_distance <- dist(standardised_data, method = "manhattan")

single_manhattan <- hclust(manhattan_distance, method = "single")
complete_manhattan <- hclust(manhattan_distance, method = "complete")
average_manhattan <- hclust(manhattan_distance, method = "average")

par(mfrow = c(1, 3))
plot(single_manhattan, main = "Single linkage", sub = "Manhattan",
     xlab = "", cex = 0.6)
plot(complete_manhattan, main = "Complete linkage", sub = "Manhattan",
     xlab = "", cex = 0.6)
plot(average_manhattan, main = "Average linkage", sub = "Manhattan",
     xlab = "", cex = 0.6)
par(mfrow = c(1, 1))

cluster_complete_manhattan <- cutree(complete_manhattan, k = 3)
table(
  Cluster = cluster_complete_manhattan,
  Diagnosis = prostate$diagnosis_result
)
```

## Hierarchical clustering with Euclidean distance

The Euclidean comparison retains single, complete, average, centroid, and Ward.D2 linkage. `ward.D2` is used for the minimum-variance solution.

```{r}
euclidean_distance <- dist(standardised_data, method = "euclidean")

single_euclidean <- hclust(euclidean_distance, method = "single")
complete_euclidean <- hclust(euclidean_distance, method = "complete")
average_euclidean <- hclust(euclidean_distance, method = "average")
centroid_euclidean <- hclust(euclidean_distance, method = "centroid")
ward_euclidean <- hclust(euclidean_distance, method = "ward.D2")

par(mfrow = c(2, 3))
plot(single_euclidean, main = "Single linkage", sub = "Euclidean",
     xlab = "", cex = 0.6)
plot(complete_euclidean, main = "Complete linkage", sub = "Euclidean",
     xlab = "", cex = 0.6)
plot(average_euclidean, main = "Average linkage", sub = "Euclidean",
     xlab = "", cex = 0.6)
plot(centroid_euclidean, main = "Centroid linkage", sub = "Euclidean",
     xlab = "", cex = 0.6)
plot(ward_euclidean, main = "Ward.D2", sub = "Euclidean",
     xlab = "", cex = 0.6)
plot.new()
par(mfrow = c(1, 1))

cluster_complete_euclidean <- cutree(complete_euclidean, k = 3)
cluster_ward_euclidean <- cutree(ward_euclidean, k = 3)

table(
  Cluster = cluster_ward_euclidean,
  Diagnosis = prostate$diagnosis_result
)
```

## Cluster profiles for hierarchical solutions

```{r}
plot_cluster_boxplots <- function(groups, title_prefix) {
  par(mfrow = c(2, 4))

  for (i in seq_along(clustering_variables)) {
    boxplot(
      original_scale_data[[i]] ~ factor(groups),
      col = rainbow(3),
      main = paste(title_prefix, clustering_variables[i]),
      xlab = "Cluster",
      ylab = ""
    )
  }

  par(mfrow = c(1, 1))
}

plot_cluster_boxplots(
  cluster_complete_manhattan,
  "Complete Manhattan:"
)
plot_cluster_boxplots(
  cluster_ward_euclidean,
  "Ward.D2:"
)
```

## K-means clustering

The original three-cluster analysis is retained. `set.seed()` and multiple starts make the solution reproducible and reduce dependence on one random initialization.

```{r}
set.seed(125)
fviz_nbclust(
  standardised_data,
  kmeans,
  method = "wss",
  k.max = 10,
  nstart = 100
)
```

```{r}
set.seed(125)
kmeans_model <- kmeans(
  standardised_data,
  centers = 3,
  iter.max = 100,
  nstart = 100
)

table(kmeans_model$cluster)
table(
  Cluster = kmeans_model$cluster,
  Diagnosis = prostate$diagnosis_result
)
kmeans_model$centers
```

```{r}
plot_cluster_boxplots(kmeans_model$cluster, "K-means:")
```

## Gap statistic

```{r}
set.seed(125)
gap_statistic <- clusGap(
  standardised_data,
  FUNcluster = kmeans,
  K.max = 8,
  B = 100,
  nstart = 50
)

fviz_gap_stat(gap_statistic)
```

## Model-based clustering

```{r}
set.seed(125)
model_based <- Mclust(standardised_data, G = 2:6, verbose = FALSE)

plot(model_based, what = "BIC")
summary(model_based)
table(model_based$classification)
table(
  Cluster = model_based$classification,
  Diagnosis = prostate$diagnosis_result
)
```

## Principal component analysis

PCA uses the already standardised variables, so no second standardisation is applied.

```{r}
pca_model <- princomp(standardised_data, cor = FALSE)
summary(pca_model)
```

```{r}
plot_clusters_on_pca <- function(cluster, title) {
  plot(
    pca_model$scores[, 1],
    pca_model$scores[, 2],
    type = "n",
    xlab = "First principal component",
    ylab = "Second principal component",
    main = title
  )
  text(
    pca_model$scores[, 1],
    pca_model$scores[, 2],
    labels = cluster,
    col = cluster
  )
}

par(mfrow = c(3, 2))
plot_clusters_on_pca(
  cluster_complete_manhattan,
  "Complete linkage: Manhattan"
)
plot_clusters_on_pca(
  cluster_ward_euclidean,
  "Ward.D2: Euclidean"
)
plot_clusters_on_pca(
  cluster_complete_euclidean,
  "Complete linkage: Euclidean"
)
plot_clusters_on_pca(kmeans_model$cluster, "K-means")
plot_clusters_on_pca(model_based$classification, "Model-based")
plot(
  pca_model$scores[, 1],
  pca_model$scores[, 2],
  col = prostate$diagnosis_result,
  pch = 19,
  xlab = "First principal component",
  ylab = "Second principal component",
  main = "Reference diagnosis"
)
par(mfrow = c(1, 1))
```

```{r}
biplot(pca_model, choices = c(1, 2), xlabs = prostate$id, cex = 0.6)
```

```{r}
loadings <- pca_model$loadings[, 1:4]
loadings_data <- as.data.frame(loadings)
loadings_data$Variable <- rownames(loadings_data)
loadings_long <- melt(loadings_data, id.vars = "Variable")

ggplot(
  loadings_long,
  aes(x = Variable, y = value, fill = variable)
) +
  geom_col(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "PCA loadings",
    x = "Variable",
    y = "Loading",
    fill = "Component"
  )
```

## Internal evaluation

`eclust()` receives `k = 3` explicitly. Silhouette widths are calculated with the distance appropriate to each method.

```{r}
set.seed(125)
eclust_kmeans <- eclust(
  standardised_data,
  FUNcluster = "kmeans",
  k = 3,
  nstart = 100,
  graph = FALSE
)

eclust_ward_euclidean <- eclust(
  standardised_data,
  FUNcluster = "hclust",
  k = 3,
  hc_metric = "euclidean",
  hc_method = "ward.D2",
  graph = FALSE
)

eclust_complete_manhattan <- eclust(
  standardised_data,
  FUNcluster = "hclust",
  k = 3,
  hc_metric = "manhattan",
  hc_method = "complete",
  graph = FALSE
)

model_based_silhouette <- silhouette(
  model_based$classification,
  euclidean_distance
)

internal_evaluation <- data.frame(
  Method = c(
    "K-means",
    "Model-based",
    "Ward.D2 Euclidean",
    "Complete Manhattan"
  ),
  Average_Silhouette = c(
    eclust_kmeans$silinfo$avg.width,
    mean(model_based_silhouette[, "sil_width"]),
    eclust_ward_euclidean$silinfo$avg.width,
    eclust_complete_manhattan$silinfo$avg.width
  )
)

internal_evaluation[order(-internal_evaluation$Average_Silhouette), ]
```

## Agreement between clustering solutions

Adjusted Rand Index and normalized mutual information are invariant to cluster-label permutations.

```{r}
normalised_mutual_information <- function(first_partition,
                                          second_partition) {
  contingency <- table(first_partition, second_partition)
  joint_probability <- contingency / sum(contingency)
  first_probability <- rowSums(joint_probability)
  second_probability <- colSums(joint_probability)

  row_probability <- first_probability[row(joint_probability)]
  column_probability <- second_probability[col(joint_probability)]
  nonzero <- joint_probability > 0

  mutual_information <- sum(
    joint_probability[nonzero] *
      log(
        joint_probability[nonzero] /
          (row_probability[nonzero] * column_probability[nonzero])
      )
  )

  first_entropy <- -sum(
    first_probability[first_probability > 0] *
      log(first_probability[first_probability > 0])
  )
  second_entropy <- -sum(
    second_probability[second_probability > 0] *
      log(second_probability[second_probability > 0])
  )

  mutual_information / sqrt(first_entropy * second_entropy)
}

method_comparison <- data.frame(
  Comparison = c(
    "Ward.D2 Euclidean vs K-means",
    "Complete Manhattan vs K-means",
    "Model-based vs K-means"
  ),
  Adjusted_Rand_Index = c(
    adjustedRandIndex(cluster_ward_euclidean, kmeans_model$cluster),
    adjustedRandIndex(cluster_complete_manhattan, kmeans_model$cluster),
    adjustedRandIndex(model_based$classification, kmeans_model$cluster)
  ),
  Normalised_Mutual_Information = c(
    normalised_mutual_information(
      cluster_ward_euclidean,
      kmeans_model$cluster
    ),
    normalised_mutual_information(
      cluster_complete_manhattan,
      kmeans_model$cluster
    ),
    normalised_mutual_information(
      model_based$classification,
      kmeans_model$cluster
    )
  )
)

method_comparison
```

## External comparison with diagnosis

Diagnosis is not used to construct clusters. The following values measure agreement only; they are not supervised classification error rates.

```{r}
external_evaluation <- data.frame(
  Method = c(
    "K-means",
    "Model-based",
    "Ward.D2 Euclidean",
    "Complete Manhattan"
  ),
  Adjusted_Rand_Index = c(
    adjustedRandIndex(kmeans_model$cluster, prostate$diagnosis_result),
    adjustedRandIndex(model_based$classification, prostate$diagnosis_result),
    adjustedRandIndex(cluster_ward_euclidean, prostate$diagnosis_result),
    adjustedRandIndex(cluster_complete_manhattan, prostate$diagnosis_result)
  ),
  Normalised_Mutual_Information = c(
    normalised_mutual_information(
      kmeans_model$cluster,
      prostate$diagnosis_result
    ),
    normalised_mutual_information(
      model_based$classification,
      prostate$diagnosis_result
    ),
    normalised_mutual_information(
      cluster_ward_euclidean,
      prostate$diagnosis_result
    ),
    normalised_mutual_information(
      cluster_complete_manhattan,
      prostate$diagnosis_result
    )
  )
)

external_evaluation[order(-external_evaluation$Adjusted_Rand_Index), ]
```

## Cluster descriptions

Cluster summaries are reported on the original measurement scale. They describe relative feature levels and do not justify clinical claims about aggressiveness or prognosis.

```{r}
cluster_profiles <- original_scale_data |>
  mutate(Cluster = factor(kmeans_model$cluster)) |>
  group_by(Cluster) |>
  summarise(
    across(all_of(clustering_variables), mean),
    Observations = n(),
    .groups = "drop"
  )

cluster_profiles
```

## Sensitivity analysis: standard versus robust scaling

Potential outliers are retained. To examine their influence without mean replacement, K-means and Ward.D2 are repeated after median/IQR scaling.

```{r}
robust_scale <- function(x) {
  scale_value <- IQR(x)
  if (scale_value == 0) scale_value <- mad(x)
  (x - median(x)) / scale_value
}

robustly_scaled_data <- as.data.frame(
  lapply(original_scale_data, robust_scale)
)

set.seed(125)
kmeans_robust <- kmeans(
  robustly_scaled_data,
  centers = 3,
  iter.max = 100,
  nstart = 100
)

ward_robust <- hclust(
  dist(robustly_scaled_data, method = "euclidean"),
  method = "ward.D2"
)
cluster_ward_robust <- cutree(ward_robust, k = 3)

scaling_sensitivity <- data.frame(
  Method = c("K-means", "Ward.D2 Euclidean"),
  ARI_standard_vs_robust = c(
    adjustedRandIndex(kmeans_model$cluster, kmeans_robust$cluster),
    adjustedRandIndex(cluster_ward_euclidean, cluster_ward_robust)
  ),
  NMI_standard_vs_robust = c(
    normalised_mutual_information(
      kmeans_model$cluster,
      kmeans_robust$cluster
    ),
    normalised_mutual_information(
      cluster_ward_euclidean,
      cluster_ward_robust
    )
  )
)

scaling_sensitivity
```

## Conclusion

The original sequence—exploration, hierarchical clustering, K-means, model-based clustering, PCA, silhouette analysis, gap statistic, and comparison with diagnosis—is retained. The corrections preserve continuous measurements, avoid Ward clustering with Manhattan distance, make random initialization reproducible, create every referenced object, report computed rather than copied silhouette values, use ARI/NMI for external agreement, and assess outlier sensitivity without manufacturing observations through mean replacement.
